A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-29 latex/main.tex
Updates from Overleaf

[TODO we should mention UShER here - what do they do with dels?]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to disintinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. 
To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

>>>>

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice
way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to distinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. 
To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.
